---
title: "Statistical Inference course project: part 2"
author: "Evgeny Petrovskiy"
date: '2017-10-21'
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

In this assignment we are going to analyze ToothGrowth data 

load ToothGrowth dataset
```{r load}
data("ToothGrowth")
tg <- ToothGrowth
str(tg)
```

display data 
let's break data into groups that represent dose and then look compare supplemet types used for the same doses.

```{r plot}
gg <- ggplot(data = tg, aes(x = supp, y = len, color = supp)) +
  #geom_point(shape = 1) + 
  geom_boxplot() +
  #geom_jitter() +
  facet_grid(. ~ dose)
gg
```

summary of data

```{r summary}
summary(tg)
```

get len values for each dose and supplement type

```{r}
vc05 <- subset(tg, tg$sup == "VC" & tg$dose == 0.5)$len
vc10 <- subset(tg, tg$sup == "VC" & tg$dose == 1.0)$len
vc20 <- subset(tg, tg$sup == "VC" & tg$dose == 2.0)$len
oj05 <- subset(tg, tg$sup == "OJ" & tg$dose == 0.5)$len
oj10 <- subset(tg, tg$sup == "OJ" & tg$dose == 1.0)$len
oj20 <- subset(tg, tg$sup == "OJ" & tg$dose == 2.0)$len
```

## test 1: relation between dose and length

let's say that 1.0 mg of vitamin C received as ascorbic acid gives bigger effect on length than 0.5 mg of the same supplement type. Null hypotesis is that length doesn't depend on dose. Alternative hypothesis is that average value for 1.0 mg is larger than average value for 0.5 Dataset includes only 10 measurements for each type for each dose so we should use t-test. All observations are independent, paired = fasle

```{r "t-test vc 05 vs vc10"}
t.test(vc10, vc05, paired = FALSE, alternative = "g")
```

result shows that confidence interval doesn't includes 0, which means that alternative hypothesis is true. The 1.0 mg of vitamin C received as ascorbic acid results in bigger length.

## test 2: difference in length for different supplement types
Say that oj05 is more effective than vc05
null hypotesys ia that orange juice and vitamin C are equally effective (result in the same length) in 5mg dose. Alternative hypothesis is that they have different impact on growth.

```{r "t-test vc 05 vs oj05"}
t.test(oj05, vc05, paired = FALSE)
```

Lastly hypothesis that 2mg of vitamin C received in orange juice and the same dose received as ascorbic acid have different length

```{r "t-test vc 05 vs vc10"}
t.test(oj20, vc20, paired = FALSE)
```

for this test 95% confidence interval includes 0 so we failed to reject null hypothesis.
