---
title: "Statistical Inference course project"
author: "Evgeny Petrovskiy"
date: "10/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Defining parameters

In this project we will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. we will investigate the distribution of averages of 40 exponentials doing a thousand simulations.

## Exploring Exponential distribution

According to assignment we need to run 1000 simulations to find averages of 40 exponentials within each simulation. Set parameters up

```{r set parameters}
# make result reproducible
set.seed(123)
# number of simulations
nsim <- 1000
# exp distribution parameters
n <- 40; lambda <- 0.2
```

Run simulations and before we calculate averages let's look into original distribution
```{r run-sim}
simulation <- rexp(n * nsim, lambda)
hist(simulation)
```
and calculate avewrage for each simulation
```{r calc-avg}
dist <- apply(matrix(simulation, nrow = nsim), 1, mean)
```

Look at a histogram of this distribution and at its summary
```{r histogram}
hist(dist)
summary(dist)
```

According to [Wikipedia][https://en.wikipedia.org/wiki/Exponential_distribution#Mean.2C_variance.2C_moments_and_median] 
* theoretical mean E[X] = 1 / lambda = beta
* theoretical variance Var[X] = 1 / lambda^2 = beta^2

Knowing that we can calculate theoretical values
```{r theoretocal-vals}
#calculate results
tmean <- 1 / lambda; tvar <- beta^2
```
And compare them to distribution statistics. Means
```{r means-compare}
means <- c(tmean, mean(dist))
names(means) <- c("Theoretical Mean", "Distribution Mean")
means
```
As we can see distribution mean is very close to theoretical mean, this is exactly what LLN (law of large numbers) states: average limits to what it estimating, the population mean.
```{r plot-means}
ggplot(df, aes(m)) + geom_histogram(bins = 20) + geom_vline(xintercept = tmean, color = "red") + geom_vline(xintercept = mean(dist), color = "blue")
```
On the plot above we see red line that represents theoretical meand and blue line that shows mean of distribution. Lines are very close to each other due to big number of simulations. 

Variances
```{r vars-compare}
variances <- c(tvar, var(dist))
names(variances) <- c("Theoretical Var", "Distribution Var")
variances
```
These two values are complete diffirent because variance of means of samples has nothing common with theoretical variance of exponential distribution.

Let's look at disctributions. For this task we will find statistics with t-test
```{r t.test}
tt <- t.test(dist, mu = tmean, conf.level = 0.95)
tt$estimate

tt$conf.int
```

now we will plot density of our distribution and highlight confidence intervals
```{r distr}
df <- data.frame(m = dist)
g <- ggplot(data = df, aes(m)) + 
  geom_density() +
  geom_vline(xintercept = tt$conf.int)
g
```