---
title: "Statistical Inference course project"
author: "Evgeny Petrovskiy"
date: "10/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Defining parameters

In this project we will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. we will investigate the distribution of averages of 40 exponentials doing a thousand simulations.

## Exploring Exponential distribution

```{r set parameters}
set.seed(123)
n <- 40
lambda <- 0.2
dist <- rexp(n, lambda)
```

Let's have a look at a histogram and a summary of this distribution
```{r summary}
hist(dist)
```

According to [Wikipedia][https://en.wikipedia.org/wiki/Exponential_distribution#Mean.2C_variance.2C_moments_and_median] 
* theoretical mean E[X] = 1 / lambda = beta
* theoretical variance Var[X] = 1 / lambda^2 = beta^2

Knowing that we can calculate theoretical values
```{r theoretocal-vals}
#calculate results
tmean <- 1 / lambda; tvar <- beta^2
```
And compare them to distribution statistics. Means
```{r means-compare}
means <- c(tmean, mean(dist))
names(means) <- c("Theoretical Mean", "Distribution Mean")
means
```
Variances
```{r vars-compare}
variances <- c(tvar, var(dist))
names(variances) <- c("Theoretical Var", "Distribution Var")
variances
```

## Simulations

```{r simulation}
nsim <- 1000
sdist <- sample(dist, )