---
title: "Severe weather events"
output: html_document
---

```{r "initialization", include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# loading libraries
library(ggplot2)
library(dplyr)

# initializing constatns
dataurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
locfile <- "../workdata/input/StormData.csv.bz2"

```

## Synopsis

max 10 sentences {r dataurl}
what is exercise, what data is used. what are main questions to answer. how it was done. what are results.

## Data Processing

Data is the form of a comma-separated-value [file](`r dataurl`) compressed via the bzip2 algorithm to reduce its size. We download it and store in working folder for further use and loading into R. 

```{r "load data", cache=TRUE}
if ( !file.exists(locfile)) {
  download.file(dataurl, locfile, method = "curl")
}
dataset <- read.table(locfile, header = T, sep = ",", nrows = 1e5)
```

This dataset is about 500 MB and contains 37 variables. Description is provided by National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). For our study we need only some of them: let's take State, Date, Event Type, Fatalities, Injuries, Prop

```{r "pick some attributes"}
workset <- tbl_df(
  dataset[,c("STATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG")]
)
workset$BGN_DATE <- as.Date(as.character(dataset$BGN_DATE), format = "%m/%d/%Y")
```

Sumarize data

```{r "summarize by Year and Event Type}
sumyrtp <- workset %>%
  group_by(EVTYPE, YEAR = as.numeric(format(BGN_DATE, "%Y"))) %>%
  summarize(
    INJURIES = sum(INJURIES, na.rm = T), 
    FATALITIES = sum(FATALITIES, na.rm = T), 
    PROPDMG = sum(PROPDMG, na.rm = T)
  )

```
## Results

Plot injuries and fatalities by years 

```{r "people health"}
qplot(data = sumyrtp, x = YEAR, y = INJURIES, col = EVTYPE, geom = "line")
```


