---
title: "Severe weather events"
output: html_document
---

```{r "initialization"}
knitr::opts_chunk$set(echo = TRUE)

# initializing constatns
dataurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
locfile <- "../workdata/input/StormData.csv.bz2"

# loading libraries
library(ggplot2)
library(tidyr)
library(dplyr)

```

## Synopsis

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. Questions to be answered:

1. most harmful events with respect to population health
2. events with the greatest economic consequences

We use U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to find answer to these questions. And look a bit closer at events that impact on people health.  

## Data Processing

Data is available in the form of a comma-separated-value [file](`r dataurl`) compressed via the bzip2 algorithm to reduce its size. We download it and store in working folder for further use in our study. 

```{r "load data", cache = TRUE}
if ( !file.exists(locfile)) {
  download.file(dataurl, locfile, method = "curl")
}
dataset <- read.table(locfile, header = T, sep = ",")
```

This dataset is about 500 MB and contains 37 variables. Description is provided by National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf). For our study we need only some of them: let's take State, Date, Event Type, Fatalities, Injuries, PropDmg, CropDmg. 

We will use

* `FATALITIES` and `INJURIES` to assess impact on health;
* `PROPDMG` and `CROPDMG` to assess economic consequences.

```{r "pick attributes", cache = TRUE}
# extract necessary columns
workset <- tbl_df(
  dataset[,c("STATE", "EVTYPE", 
             "FATALITIES", "INJURIES", 
             "PROPDMG", "CROPDMG", 
             "CROPDMGEXP", "PROPDMGEXP",
             "BGN_DATE"
             )]
  )
```

In order to manupulate with Damage values they should be *normalized* first. Based on respective EXP columns that contain `K` (thousands), `M` (millions), `B` (billions) we can come to common demoninator. We also translate date into date format.

```{r normalize, cache = TRUE}
workset <- workset %>%
  # convert date and amount orders (K = 10^3, M - 10^6, B = 10^9)
  mutate(
    BGN_DATE = as.Date(as.character(BGN_DATE), format = "%m/%d/%Y"),
    PROPDMGEXP = toupper(PROPDMGEXP),
    CROPDMGEXP = toupper(CROPDMGEXP)
  ) %>%
  left_join(
    tbl_df(data.frame(PROPDMGEXP = c('K', 'M', 'B'), prop.ratio = c(1e3, 1e6, 1e9)))
  ) %>%
  left_join(
    tbl_df(data.frame(CROPDMGEXP = c('K', 'M', 'B'), crop.ratio = c(1e3, 1e6, 1e9)))
  ) %>%
  # translate damage into USD
  mutate(
    PROPERTYDAMAGE = PROPDMG * ifelse(is.na(prop.ratio), 1, prop.ratio),
    CROPDAMAGE = CROPDMG * ifelse(is.na(crop.ratio), 1, crop.ratio)
  ) 
```

We gather variables `FATALITIES`, `INJURIES`, `PROPDMG` and `CROPDMG` into one column because it becomes easier to make comparative analysis. We group data by years because information about exact date is excessive in our case.

```{r melt, cache = TRUE}
workset <- workset %>%
  # make it long
  gather(
    key = "IMPACT", 
    value = "VALUE", 
    c(INJURIES, FATALITIES, PROPERTYDAMAGE, CROPDAMAGE)
  ) %>%
  mutate(
    IMPACTTYPE = ifelse(IMPACT %in% c("INJURIES", "FATALITIES"), "HEALTH", "ECONOMY")
  ) %>%
  # keep only useful columns
  select(
    EVTYPE, IMPACTTYPE, IMPACT, VALUE, BGN_DATE
  ) %>%
  group_by(
    EVTYPE, IMPACTTYPE, IMPACT, 
    YEAR = as.numeric(format(BGN_DATE, "%Y"))
  ) %>%
  summarise(
    VALUE = sum(VALUE, na.rm = T)
  )
```

To find most harmful events we compare total impact over all years. 

Sumarize data: group data by impact and event types. Rank event types by impact; top-5 event types with highest impact will be plotted later.

```{r "top5 selection"}
top5ev <- workset %>%
# group by IMPACT and calc totals
  group_by(
    EVTYPE,
    IMPACTTYPE,
    IMPACT
  ) %>%
  summarise(
    VALUE = sum(VALUE, na.rm = T)
  ) %>%
  group_by(
    IMPACT
  ) %>%
  mutate(
    place = row_number(desc(VALUE))
  ) %>%
  filter(
    place <= 5
  ) %>%
  mutate(
    graphorder = 6 - place
  )
```

For event types that impact health most (#1 in top-5 ranking) take summary by years of observation. 

```{r "summarize by Year and Event Type"}
sumyrtp <- workset %>%
  group_by( 
    EVTYPE, 
    YEAR, 
    IMPACTTYPE,
    IMPACT 
  ) %>%
  summarize( 
    VALUE = sum(VALUE, na.rm = T) 
  ) %>%
  inner_join(
    subset(top5ev[c("EVTYPE", "IMPACT", "place")], place == 1),
    by = c("EVTYPE" = "EVTYPE", "IMPACT" = "IMPACT")
  )
```


## Results

### Property damage by event types

Let's plot property damage by event types

```{r "plot prop damage"}
ggplot(
  subset(top5ev, IMPACTTYPE == "ECONOMY"),
  aes(x=reorder(EVTYPE, graphorder), y=VALUE/1e9, fill = EVTYPE)
  ) +
  geom_bar(stat='identity', show.legend = F, width = .75) +
  facet_grid(facets = IMPACT ~ .) +
  labs(
    title = "Property damage by Event Types", 
    x = "Event Type", y = "Damage in Billions of US Dollars"
  ) +
  coord_flip()
```

As it can bee seen, DROUGH has biggest negatinve impact to agriculture while it doesn't harm property. And FLOOD is most harmful weather event for property while it is not that severe for crop. Another conclusion is that property is more exposed to severe weather events compare to crop. At least in our Top-5 rating.

### Injuries and fatalities by event types

Let's plot how weather events types affect people health and find out top-5 with highest impact.

```{r "plot people health"}
ggplot(
  subset(top5ev, IMPACTTYPE == "HEALTH"),
  aes(x=reorder(EVTYPE, graphorder), y=VALUE/1e3, fill = EVTYPE)
  ) +
  geom_bar(stat='identity', show.legend = F, width = .75) +
  facet_grid(facets = IMPACT ~ .) +
  labs(
    title = "Impact on people health over years",
    x = "Event Type", y = "Thousands of People"
  ) +
  coord_flip()
```

Chart above clearly shows that tornados have most impact on people health. Number of injuries is 10 times higher than any other event.

### History of tornado

Most harmful event type for people health is TORNADO.
Let's look what were observations for it.

```{r "plot history"}
ggplot(
  subset(sumyrtp, IMPACTTYPE == "HEALTH"),
  aes(x=YEAR, y=VALUE, fill = EVTYPE)
  ) +
  geom_bar(stat='identity', show.legend = F, width = .75) +
  facet_grid(facets = IMPACT ~ ., scales = "free_y") +
  labs(
    title = "TORNADO impact on population health",
    x = "Year", y = "Number of People"
  )
```

Number of injuries and fatalities didn't descrease in course of last years. Year 2011 observations show that number of fatalities is highest since the beginning of observations. Number of injuries is number 2 among other years.
