---
title: "Regression Models Course Project"
author: "Evgeny Petrovskiy"
date: "11/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)

```

## first look at data

Goal of this assignment is analyze `mtcars` dataset and to answer two questions

* 

Below is desription of columns for mtcars dataset: this is a data frame with 32 observations on 11 variables.

1.	mpg	Miles/(US) gallon
2.	cyl	Number of cylinders
3.	disp	Displacement (cu.in.)
4.	hp	Gross horsepower
5.	drat	Rear axle ratio
6.	wt	Weight (1000 lbs)
7.	qsec	1/4 mile time
8.	vs	V/S
9.	am	Transmission (0 = automatic, 1 = manual)
10.	gear	Number of forward gears
11.	carb	Number of carburetors


```{mtcars}
str(mtcars)
```

## Exploratory analysys

```{r "explore"}
p <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) + geom_point()
p + facet_grid(. ~ am)

p + facet_grid(. ~ am, aes(disp, mpg))
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)

names(mtcars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
