---
title: "Regression Models Course Project"
author: "Evgeny Petrovskiy"
date: "11/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)

```
## Summary

Goal of this assignment is to analyze `mtcars` dataset and answer two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

In order to answe rthis questions we apply different techniques Below: first we do exploratory analysis, then we apply regression models and assess their quality, finally we are trying to answer questions based on information we gathered.

## first look at data

Below is desription of columns for mtcars dataset: this is a data frame with 32 observations on 11 variables.

1.	`mpg`	Miles/(US) gallon
2.	`cyl`	Number of cylinders
3.	`disp`	Displacement (cu.in.)
4.	`hp`	Gross horsepower
5.	`drat`	Rear axle ratio
6.	`wt`	Weight (1000 lbs)
7.	`qsec`	1/4 mile time
8.	`vs`	V/S
9.	`am`	Transmission (0 = automatic, 1 = manual)
10.	`gear`	Number of forward gears
11.	`carb`	Number of carburetors


```{r mtcars}
str(mtcars)
```

## Exploratory analysys

```{r "explore"}
plotx <- function(x) {
  plot(
    mtcars$mpg ~ mtcars[,x], 
    col = factor(mtcars$am), 
    pch = 19, 
    xlab = x, ylab = "mpg"
  )
}
cols <- names(mtcars)
cols <- cols[!cols %in% c("mpg", "am")]
par(mfrow = c(1,1))
# sapply(X = cols, FUN = plotx)
for (i in cols) plotx(i)

```

looking at data above we can make a conclusion that *mpg* value correlates with *wt*, *disp*, *hp*

let's look at them on logarirhmic scale
```{r "closelook"}
for (i in c("disp", "hp", "wt")) {
  plot(
    log(mtcars$mpg) ~ mtcars[, i],
    col = factor(mtcars$am), pch = 19, xlab = i
  )
}
```

FolLowing correlations look very strong 

* `Miles per galon` and `Displacement`
* `Miles per galon` and `Horse powers`
