---
title: "Regression Models Course Project"
author: "Evgeny Petrovskiy"
date: "11/18/2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)

```
## Summary

Goal of this assignment is to analyze `mtcars` dataset and answer two questions:

* "Is an automatic or manual transmission better for MPG"
* "Quantify the MPG difference between automatic and manual transmissions"

In order to answer this question we apply different techniques below: first we do exploratory analysis, then we apply regression models and assess their quality, finally we are trying to answer questions based on information we gathered.

## First look at data

`mtcars` is a data frame with 32 observations that include 11 different measurements. Below is desription of columns for  dataset: 

1.	`mpg`	Miles/(US) gallon
2.	`cyl`	Number of cylinders
3.	`disp`	Displacement (cu.in.)
4.	`hp`	Gross horsepower
5.	`drat`	Rear axle ratio
6.	`wt`	Weight (1000 lbs)
7.	`qsec`	1/4 mile time
8.	`vs`	V/S
9.	`am`	Transmission (0 = automatic, 1 = manual)
10.	`gear`	Number of forward gears
11.	`carb`	Number of carburetors


```{r mtcars}
str(mtcars)
```

## Exploratory analysys

We are trying to establish relation between mpg and other measured parameters. Appendix I shows relation bewteen MPG and all other variables of the dataset. Looking at that data we can make a conclusion that *mpg* value correlates with *wt*, *disp*, *hp*

let's look at them on logarirhmic scale
```{r "closelook"}
for (i in c("disp", "hp", "wt")) {
  plot(
    log(mtcars$mpg) ~ mtcars[, i],
    col = factor(mtcars$am), pch = 19, xlab = i,
    main = paste("logarithmic relation between MPG and", toupper(i), sep = " ")
  )
}
```

Following correlations look very strong 

* `Miles per galon` and `Wieght`
* `Miles per galon` and `Displacement`
* `Miles per galon` and `Gross horsepower`

Relation bewteen displacement & horse powers

```{r rel-disp-hp}
plot(hp ~ disp, data = mtcars, main = "Relation between horsepower and displacement")
```

## Fitting models

Let's build models that involve transmission type (`am`), displacement (`disp`) and horse powers (`hp`) as regressors and perform anova test. 

``` {r "models"}
fit1 <- lm(log(mpg) ~ wt, data = mtcars)
fit2 <- lm(log(mpg) ~ wt + disp, data = mtcars)
fit3 <- lm(log(mpg) ~ wt + disp + hp, data = mtcars)
fit.all <- lm(log(mpg) ~ ., data = mtcars)

anova(fit1, fit2, fit3, fit.all)
```

Anova test shows that after considering `wt` model can be slightly improved with `disp` and `hp` but other regressors doesn't make model significantly better.

```{r diagnose-fit1}
summary(fit1)

plot(
  x = mtcars$wt, y = fit1$res, xlab = "Weight", ylab = "residuals",
  main = "Residuals plot for log(MPG) ~ WT model")
abline(0, 0)
hist(fit1$res)
```

Regression model for MPG predicted based on Weight has r^2 ~ 80%, residuals are normally distributed around 0, histogram doesn't look smooth because we have few (32) observations.

Let's exclude displacement and horsepower from regressors list and compare models that involve wieght only and weight & am-characteristic.

```{r am-anova}
fit1   <- lm(log(mpg) ~ wt, data = mtcars)
fit1am <- lm(log(mpg) ~ wt + am, data = mtcars)

anova(fit1, fit1am)
```

Anova test doesn't find `am` regressor to be significant. For data in the dataset we can't say that automatic-manual transmission makes any difference in model fitting.

Before making conclusions let's separately fit models for manual and automatic transmissions and visualize them on one chart. Please be aware that chart shows `mpg` in log units (so we need to exponentiate them to get real data)

```{r am-models}
fitA <- lm(log(mpg) ~ wt, data = subset(mtcars, am == 0))
fitM <- lm(log(mpg) ~ wt, data = subset(mtcars, am == 1))

plot(log(mpg) ~ wt, col = factor(am), data = mtcars, pch = 18,
     main = "Regression lines for Auto and Manual transmissions")
#plot(exp(fitA$fitted.values) ~ subset(mtcars, am == 0)$wt)
abline(fitA, col = 1)
abline(fitM, col = 2)

```

## Conclusions

Trying to answer questions about which transmission is better first of all we have to mention that data is not sufficient enough to draw conclusion because: * other parameters for cars vary a lot (we do not have paired obsevations)
* tests regression models do not recognize `am` regressor as meaningful

But if we had to make some assumptions based on last figure "Regression lines for Auto and Manual transmissions" we would say that 

1. cars with automatic transmission better for heavy cars (weight > 3000 lbs) 
2. fuel consumtion for cars with automatic transmission increases less with weight gain

below are calculations for change in MPG for car with average weight if we would increase it by 1000 lbs.

```{r coefficients}
mwt <- mean(mtcars$wt)
print(paste("Average weight in lbs", round(mwt, 1)*1000))
deltaA <- sum(c(-1, 1) * exp(predict(fitA, data.frame(wt = c(mwt, mwt + 1)))))
deltaM <- sum(c(-1, 1) * exp(predict(fitM, data.frame(wt = c(mwt, mwt + 1)))))
print(paste("Change in MPG with 1000 lbs gain to Average weight for Manual transmission", round(deltaM, 1), sep = " "))
print(paste("Change in MPG with 1000 lbs gain to Average weight for Automatic transmission", round(deltaA, 1), sep = " "))
```

As we can see decrease in MPG is less significant for cars with automatic transmission.

# Appendices

## Appendix I

```{r "explore"}
cols <- names(mtcars)
cols <- cols[!cols %in% c("mpg")]
par(mfrow = c(1,1))
# sapply(X = cols, FUN = plotx)
for (x in cols) {
  plot(
    mtcars$mpg ~ mtcars[,x], 
    col = factor(mtcars$am), 
    pch = 19, 
    xlab = x, ylab = "mpg",
    main = paste("Relation between MPG and", toupper(x), sep = " ")
  ) 
}

```
